A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-27 paper1_tKdV/tKdV_theory.tex
NM

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. Recent controlled
laboratory experiments in weakly turbulent shallow water with abrupt
depth change (ADC) exhibit a remarkable transition from nearly Gaussian
statistics in incoming wave trains before the ADC to outgoing waves
trains after the ADC with extreme anomalous statistics with large
positive skewness of the surface height. Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media.
\end{abstract}

>>>>

\begin{abstract}
Understanding and predicting extreme events and their anomalous statistics
in complex nonlinear systems is a grand challenge in climate, material,
and neuroscience, as well as for engineering design. 
Recent laboratory experiments in weakly turbulent shallow water reveal a remarkable transition from Gaussian to anomalous behavior as surface waves cross an abrupt depth change (ADC). Downstream of the ADC, PDFs of  surface displacement exhibit strong positive skewness, which is accompanied by an elevated level of extreme events. 
Here we develop a statistical
dynamical model to explain and quantitatively predict the above anomalous
statistical behavior as experimental control parameters are varied.
The first step is to use incoming and outgoing truncated Korteweg-de
Vries (TKdV) equations matched in time at the ADC. The TKdV equation
is a Hamiltonian system which induces incoming and outgoing statistical
Gibbs invariant measures.
The statistical matching of the known nearly Gaussian incoming Gibbs
state at the ADC completely determines the predicted anomalous outgoing
Gibbs state, which can be calculated by a simple sampling algorithm, verified
by direct numerical simulations, and successfully captures key features
of the experiment. There is even an analytic formula for the anomalous
outgoing skewness. The strategy here should be useful for predicting
extreme anomalous statistical behavior in other dispersive media.
\end{abstract}
